Microlensing as a possible probe of the internal structure of quasars

### 2015-02-08 paper1/discussion.tex
released draft

\section{Discussion}

>>>>

\section{Discussion}
[Mihai's contribution]\\
In the current paper we simulate and study the resulting microlensing lightcurves of geometric crescent-shaped sources 
and compare them with the microlensing lightcurves of other simple mathematically describable source profiles. 
In order to mimic the behaviour of flux of light from the source in the proximity of a fold caustic we make use of the simple approximation described in equation (5). 
The equation would exhaustively describe the magnification map and offer a good universal aproximation for the 
particular microlensing regime that we consider. 
Namely, the shape of the caustic boundary in the proximity of the source in the respective plane can be approximated with
 a line due to reason that the local radius of curvature of caustic is orders of magnitude greater than the half-light
 radius of the studied source. 
In particular cases in which the previously mentioned approximation loses its validity the impact on the quality of the 
lightcurves is not evenly distributed. The shape of the lightcurve will be maintained. 
The datapoints corresponding to the source position before and during the overlapping of the caustic will be affected by
 smaller errors than the datapoints corresponding to later times. \\

The first two source profiles that we consider are the uniform disk and symmetric gaussian source. 
Both of them can be described by a half-light radius $r_{1/2}$ and a total unlensed light flux $S_0$. 
With the two parameters constrained the one dimensional profiles as well as the lightcurves of the two source are completely determined, since no free parameter remains.

The previous statement does not hold for a crescent source.
In the case of the crescent source there are in total five parameters: the integrated flux of the source $S_0$,
 the radii of the bright/dark disk $R_p$/$R_n$ and the displacement of the centers of the two disks on the axes perpendicular and parallel to the caustic $a$ and $b$.

Two of the parameters can be reduced by expressing the results in terms of $S_0$ and $r_{1/2}$. 
The later being determinable for any set of parameters $R_p, R_n$ and $a^2+b^2$. Moreover, one of the displacemement 
parameters $b$ has no impact on the one dimensional profile of the source that results from the projection of the source image on an axis perpendicular to the caustic.

Since the one dimensional source profile that corresponds to an axis perpendicular to the caustic contains exhaustively all the information regarding the source that can be revealed by the lightcurve, 
the value of the parameter $b$ does not effect the shape of the lighcurve.

Nevertheless the $b$ parameter is relevant to the calculation of $r_{1/2}$. 
It's qualitative effect is to decrease the value of the half-light radius when the absolute value of the parameter is increased.  
With two parameters constrained and another irrelevant to the shape of the lightcurve, two free parameters remain $R_n$ and $a$. 
Figure 4 reveals that the lighcurve of a crescent source has more visible features than the other two light-curves 
corresponding to the disc and guasssian shape. The parameters $R_n$ and $a$ have strong influences on the shape of 
the microlensed lightcurve, as can be seen in figures 5, 6 and 7. Moreover, the one dimensional source profile corresponding to the direction perpendicular to the caustic reveal four characteristic points. The overlap of each of these points with the caustic leaves visible features on the lightcurve at the corresponding instances of time. In timely order the instances correspond to the start of the overlap between the caustic and the bright disk, the start of the overlap between the caustic and the dark disk, the end of the overlap between the caustic and dark disk and finally the end of the overlap between the caustic and the bright disk.

With the different source profiles and their corresponding lightcurves studied we can change our point of view of the system to that of an observer. The observer would basically detect only the lightcurve of such as source. As described in section (Lightcurve of a crescent source) the timing of the onset and offset of the previously described periods can be used to estimate the values of the radii and one of the displacement parameters when assuming a geometric crescent, the halfwidths and one displacement parameters when assuming a constant brightness elliptical crescent and the halfwidth of the dark ellipse when assuming an elliptical gaussian crescent. All quantities can be estimated in terms of the relative velocity of the source in a direction perpendicular to the caustic.\\

The previously mentioned abstract parameters can be related to physical quantities specific to the central region of a quasar. Thus the luminous region would correspond to the bright accretion disc that surrounds the black hole. The later's gravity would cause a shadow in the bright region limited by the extent of the event horizon of the black hole. Therefore the radius of the bright disk would provide an estimate of the size of the accretion disk and the radius of the dark disc would provide and estimate of the Schwarzschild radius of the black hole. Moreover a value of the Schwarzschild radius 
can be used to estimate the mass of the black hole $M \approx \frac{\Delta t_{dark}  v_p c^2}{2G}$. In the previous 
expression the $\Delta t_{dark}$ denotes the period of overlap between the black hole shadow and fold caustic. 
$v_p$ denotes the component of the relative velocity of the source and fold which is perpendicular to the caustic.  Furthermore, a simulated image of M87 presented in \citep{2012MNRAS.421.1517D} (figure 9) has been microlensed. On the resulting lightcurve the instances corresponding to the start and end of the black hole shadow and caustic overlap were distinguishable as can be seen in figure 10.

The parameters whose values cannot be determined due to the loss of information from the directions parallel to the 
caustic could be obtained in the eventuality in which the same source crosses multiple caustics that are not parallel. 
Multiple crossing of caustics can reveal details of the one dimensional flux profile corresponding to multiple distinct 
directions which would allow the reconstruction of the two dimensional profile analogous to the process through which an image of a CT scan is obtained.

In case of a high quality lightcurve with insignificant noise and measurement errors, the parameters can be obtained 
by simply identifying the characterstic instances of time withought making use of the actual values of the magnification
 map. If the effect of the errors and the noise distorts the magnification time function enough so that the 
characteristic points are not identifiable with the characteristic periods still visible, the boundaries of the periods 
can be roughly estimated. To study if good estimates of the parameters can still be obtained when datapoints are
 strongly affected by the noise we propose ....

[Irshad's contribution]\\

### 2015-02-08 paper1/theory.tex
Intro polishing
released draft

\begin{figure}
\centering
  \includegraphics[width=0.95\hsize]{plots/IRIS567_path_2.eps}
\caption{\label{fig:magnification} Magnification map}
\end{figure}

@@

@@

Due to the nature of the caustic and the monotonic behaviour of the magnification map on both sides of the caustic. , the previously mentioned characteristic points are inherited by the microlensing lightcurve. The points on the temporal dimension $t_1, t_2, t_3$ and $t_4$ corresnpond to instances in time when the fold is aligned with $p_1, p_2, p_3$ and $p_4$, respectively. For a constant relative velocity $v_p$ between the source and the caustic there is a simple relation between the points $p_i$ and instances $t_i$:

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/ch_points.eps}
        \includegraphics[width = 0.48\textwidth]{plots/ch_instances.eps}
\caption{\label{fig:char_points} The characteristic points and instances of a crescent source's one dimensional profile and microlensing lightcurve.}
\end{figure}

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/infinite_fold_ellipsess.eps}
\caption{\label{fig:ellipse_profile} Elliptical sources with constant and gaussian distributed brightness.}
\end{figure}

\subsection{Extension to ellipticaly shaped crescent sources}

We further consider a source built from the superposition of two unalligned ellipses of constant surface brightness. Analogous to the previously presented crescent source the dark ellipse is interior to the bright ellipse and the distance between their centers projected on the lines perpendicular and parallel to the fold are $a$ and $b$. The semidimensions are ($a_p, b_p$) for thebright ellipse and ($a_n, b_n$) for the dark ellipse. The semimajor axis of the ellipses for the angles $\delta_p$ and $\delta_n$ with the caustic resulting in projected widths $2 h_p$ and $2h_n$ of the two ellipses on the axis perpendicular to the caustic.

The two dimensional description of the source is:

\begin{equation}
S_{2D}^{CEC} = S_{2D}^{CEP} - S_{2D}^{CEN}
\end{equation}

Where

\begin {equation}
S_{2D}^{CEP} = \frac{S_0^{CEC}}{2 \pi \left(a_p b_p - a_n b_n \right)} \Theta \left( 1 - \frac{\left( \left( p -p_{sp} \right) \sin{\delta_p} + \left( q - q_{sp} \right) \cos{\delta_p} \right)^2}{a_p^2}  - \frac{\left( \left( p -p_{sp} \right) \cos{\delta_p} + \left( q - q_{sp} \right) \sin{\delta_p} \right)^2}{ b_p^2} \right), 
\end {equation}
and
\begin {equation}
S_{2D}^{CEN} = \frac{S_0^{CEC}}{2 \pi \left(a_p b_p - a_n b_n \right)} \Theta \left( 1 - \frac{\left( \left( p -p_{sn} \right) \sin{\delta_n} + \left( q - q_{sn} \right) \cos{\delta_n} \right)^2}{a_n^2}  - \frac{\left( \left( p -p_{sn} \right) \cos{\delta_n} + \left( q - q_{sn} \right) \sin{\delta_n} \right)^2}{ b_n^2} \right), 
\end {equation}.

Analogous to the previous three equations the following relations can be obtained:

\begin{equation}
        h_p = \frac{v_p \left( t_4 -t_1 \right)}{2}, 
\end{equation}

\begin{equation}
        h_n = \frac{v_p \left( t_3 -t_2 \right)}{2}, 
\end{equation}

The time instances used in the previous equations are analogous to the ones considered in the crescent source formed from the superposition of constant surface brightness disks.

Moreover we consider a source described by an elliptical gaussian brightness function shadowed by a dark ellipse. In this case the two dimensional description of the source is given by:

\begin {equation}
S_{2D}^{GEC} = \frac{S_0^{GEC}}{2 \pi a_p b_p} \exp \left( - \frac{\left( \left( p -p_{sp} \right) \sin{\delta_p} + \left( q - q_{sp} \right) \cos{\delta_p} \right)^2}{2 a_p^2}  - \frac{\left( \left( p -p_{sp} \right) \cos{\delta_p} + \left( q - q_{sp} \right) \sin{\delta_p} \right)^2}{2 b_p^2} \right) \Pi \left(p,q\right)
\end {equation}

where the function
\begin{equation}
\Pi \left(p,q\right) = 1 - \Theta \left( 1 - \frac{\left( \left( p -p_{sn} \right) \sin{\delta_n} + \left( q - q_{sn} \right) \cos{\delta_n} \right)^2}{2 a_n^2}  - \frac{\left( \left( p -p_{sn} \right) \cos{\delta_n} + \left( q - q_{sn} \right) \sin{\delta_n} \right)^2}{2 b_n^2} \right). 
\end{equation}

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/ellipse_profile.eps}
        \includegraphics[width = 0.48\textwidth]{plots/ellipse_lightcurve.eps}
\caption{\label{fig:ellipse_profile} The one dimensional profile and resulting lightcurves of the elliptical sources.}
\end{figure}

>>>>

\begin{figure}
\centering
  \includegraphics[width=0.95\hsize]{plots/IRIS567_path_2.eps}
\caption{\label{fig:magnification_map} Magnification map}
\end{figure}

@@

\begin{figure}
\includegraphics[width = .8\textwidth]{plots/infinite_fold.eps}
\caption{\label{fig:infinite_fold}  Source profiles for gaussian, disk, crescents (right) and magnification map for an infinite fold (left). Objects have the same $S_{0}$ and $r_{1/2}$. }
\end{figure}

@@

Due to the nature of the caustic and the monotonic behaviour of the magnification map on both sides of the caustic. The previously mentioned characteristic points are inherited by the microlensing lightcurve. The points on the temporal dimension $t_1, t_2, t_3$ and $t_4$ correspond to instances in time when the fold is aligned with $p_1, p_2, p_3$ and $p_4$, respectively. For a constant relative velocity $v_p$ between the source and the caustic there is a simple relation between the points $p_i$ and instances $t_i$:

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/ch_points.eps}
        \includegraphics[width = 0.48\textwidth]{plots/ch_instances.eps}
\caption{\label{fig:char_points} The characteristic points and instances of a crescent source's one dimensional profile and microlensing lightcurve. }
\end{figure}

Figure 4 reveals that the lighcurve of a crescent source has more visible features than the other two light-curves
corresponding to the disc and guasssian shape. There are three regions where $S_{1D}$ and by inheritance the lighcurve
 have distinguishable behaviour. The first region that would be recorded on a lightcurve plot represents the preiod
of time when the bright disk begins to be overlapped by the caustic and stops when just before the dark disk reaches
the caustic. During this period of time the flux of light from the source is increasingly magnified. The second
period starts and ends with the overlapping of the dark disc. As a boundary of the two regions there is a
distinguishable point where slope of the magnification is drastically changed. This apparent discontinuity in the
first derivative of the magnification function is caused by the caustic amplification of the sudden drop in the $S_{1d}$
function. During the respective period the magnification growth slows or even reverses during the first part of the
period and starts to grow faster again as the dark disk ends its overlap with the caustic. At the point where
the dark disc clears the caustic the growth of the magnification is infinite, which appears as a saddle point
on the lightcurve. Next, the final period correspond to the case when the dark disc has cleared the caustic
and the bright disc continues to overlap with the caustic. During this period the lighcurve reaches a peak
that for most of the parameter space is global and for the rest of the parameter space local.
At the end of the period the growth of the magnification is negative infinite. the respective point appears 
as a second saddle point on the lightcurve. Past this point the magnification of any finite source will 
decay in roughly the same manner. \\

The impact of parameter $a$ on the shape of the lightcurve can be observed in figure 5 (a-var). Sources where 
the center of the dark disc reaches the caustic before the center of the bright disc are characterized by a smoother
broad peak in contrast to the cases where the center of the dark disc reaches the caustic after the brigh disk.
In the case of the latter the instance when the dark disc reaches the caustic corresponds to larger and
larger magnifications untill it becomes a local and even global peak. The effect of the $R_n$  parameter on the
lighcurve is presented in figures 6 and 7. For the particular set of parameters where $R_p = R_n +a$ the third
period of time discussed previously does not exist. Particularizing further, if the value of the radius of the dark
disc is comparable to the value of the radius of the bright one then the position and shape of the maximum magnification
 are strikingly different. In case the crescent reaches the caustic with the bright region first, the peak magnification
 happens when the dark region reaches the caustic and it is characterized by a sharp variation in magnification growth,
In the opposite case the peak appears before the end of the bright disc reaches the caustic and shape is smoother. \\

\subsection{Microlensing a simulated image of M87}

\begin{figure}
\centering
        \includegraphics[width = 0.48\textwidth]{plots/M87_image.eps}
\caption{\label{fig:M87_image} The simulated image of M87's black hole sillouette obtained by \citep{2012MNRAS.421.1517D} from a GRMHD simulation presented in \citep{2009MNRAS.394L.126M}. The color bar is linear and scaled to reach value 1 for the maximum brightness. Points p2 and p3, mark the start and end of the equivalent dark disk and are associated to the moment of their overlap with the caustic t2 and t3.}
\end{figure}

\citep{2012MNRAS.421.1517D} have created a radiative image of M87 based on the GRMHD simulations presented in \citep{2009MNRAS.394L.126M}. The respective image, reproduced in figure \ref{fig:M87_image}, has been projected to a 1D profile associated to the perpendicular direction to a fold caustic approaching the image from the right. The projection is presented in the left panel of figure 10. The amplification values of the flux of light corresponding to a microlensing event are presented in the right panel of figure 10. In general the behaviour of the lightcurve is similar to the geometric crescent source with the caveat that the outer regions surrounding the luminous parts of the image have nonzero flux and thus are more extended than the simplified source model.

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth]{plots/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained from the numerical integration along the ordonata axis of the source image (figure 9) is presented in the left subplot. The corresponding lightcurve associated to the 1D profile is presented in the right subplot. The points p2 and p3 mark the start and end of the equivalent dark disk and are associated to the moments of their overlap with the caustic: t2 and t3.}
\end{figure}
